Light inextensible strings under tension in the Schwarzschild geometry

### 2019-05-28 string_jpam.tex
use units in which G=1, basically by replacing GM with M throughout

\begin{abstract}
Here I analyze the behaviour of stationary, light inextensible strings
under tension in the vicinity of a nonrotating black hole.  Such
strings minimize their proper length, given by integrating the spatial
component of the Schwarzschild metric along the string.  The path is
calculated by numerical solution to the Euler-Lagrange equations.  If
the string is entirely outside the event horizon, its closest approach
to the singularity is tangential; if the distance of closest
approach is less than about $1.0761\cdot 2GM$, the
string self-intersects.  If the string attains $r=2GM$, then it
will in general be tangential to the event horizon.  Light taut
strings interpret the concept ``straight line'' and their behaviour
suggests a possible way of manipulating a black hole.
\end{abstract}
% AJP requires an abstract for all regular article submissions.
% Abstracts are optional for submissions to the "Notes and Discussions" section.

@@

\label{intro}
Light inextensible strings are a stalwart feature of elementary
physics.  A light string is not subject to gravity, and being under
tension means that its configuration is an extremum of proper length.
Of course, no real string is perfectly weightless or inextensible.
With tension~$T$, gravitational acceleration $g$ and (proper) mass per
(proper) unit length~$\mu$, then a string will behave as though light
on lengthscales $L\lesssim T/\left(\mu g\right)$.  Using Newtonian
expressions for the strength of gravity at the event horizon of a
black hole gives $L\lesssim TGM/\left(\mu c^2\right)$.  For the string
to behave as though light on lengthscales comparable with the
Schwarzschild radius would require $T\gtrsim\mu c^2$.  Exceeding this
criterion would violate the dominant energy condition (and in this
context we note that transverse waves would propagate at
speed~$\sqrt{T/\mu}\sim c$).  Nevertheless, the idealization of a
string as light and inextensible is an interesting and informative
limit: its analysis interprets the concept ``straight line'' and
suggests a possible way of manipulating a black hole.  Peculiar
counterintuitive behaviour such as inward-directed centrifugal force
in the vicinity of a black hole is well
known~\citep{abramowicz1992,abramowicz1993}, and the analysis of taut
strings furnishes another everyday window into general relativity.
Another possibility, presented by \citet{damour1978} and discussed by
\citet{thorne1986}, is to pass electric current through such taut
strings, an option not pursued here.

@@

\begin{equation}\label{schwarzschild}
ds^2= -dt^2\left(1-2GM/r\right) +\frac{dr^2}{1-2GM/r} + r^2\left(d\theta^2 + \sin^2\theta d\phi^2\right),
\end{equation}

\noindent where units in which~$c=1$ are used, has been known for
over a century~\cite{schwarzschild1916}.  Here I consider the spatial
component~$ds^2= dr^2/\left(1-2GM/r\right) + r^2d\phi^2$, confined to
equatorial points~($\theta=\pi/2$); we understand~$2GM\leq r<\infty$
throughout.  A circular string at constant~$r$ will, trivially, have
proper length~$2\pi r$.

Considering first the radial component we may calculate the
proper length $L\left(r_1,r_2\right)$ of a string stretching
from~$r=r_1\geq 2GM$ to~$r_2\geq r_1$ at constant $\theta,\phi$:

@@

\begin{equation}\label{radial_string_length}
  \int_{r=r_1}^{r_2}\frac{dr}{\sqrt{1-2GM/r}}=
  \left.
  \sqrt{r(r-2GM)} +\frac{1}{2}\log\left(
  \frac{\sqrt{r}+\sqrt{r-2GM}}{\sqrt{r}-\sqrt{r-2GM}}\right)
  \right|_{r_1}^{r_2}
  \end{equation}

Observe that integration may be performed with a lower limit
of~$r_1=2GM$ with no difficulty, even though the radial component of the
Schwarzschild metric approaches infinity there.

Elementary methods show that the proper length $\ell$ of radial strings has
some peculiar features.  Firstly, consider the proper length from
$r=2GM$ to $r=2GM(1+w)$ where $w>0, \left|w\right|\ll 1$.  Then

@@

 \begin{equation}\label{ell}
   \ell =
   \int_{r=2GM}^{r=2GM(1+w)}ds
   =2GM\left(2\sqrt{w} + \frac{1}{3}w^{3/2} + O(w^{5/2})\right).
 \end{equation}

This would indicate that, close to the Schwarzschild radius $r=2GM$,
large (but finite) amounts of string are required to connect two
points with slightly differing radial coordinates.  Passing to the
case $w\gg 1$ we have an asymptotic expansion that begins

% \begin{equation}\label{asymptotic_ell_properunits}
%   \ell = w +  \frac{1}{2}\log w + \frac{1}{2} + \log(2)  +  \frac{1}{8}w^{-1} + O(w^{-2}),
% \end{equation}
% 
\begin{equation}\label{asymptotic_ell}
  \ell = 2GM\left(w +  \frac{1}{2}\log w + \frac{1}{2} + \log(2)  +  \frac{1}{8}w^{-1} + O(w^{-2})\right),
\end{equation}

\noindent suggesting that there is ``extra radial space'', of length
over $1.19\cdot 2GM$, sensible at large distances from the event
horizon.  Radial strings have proper length longer than the
Schwarzschild radial coordinate difference $w$.

@@

\begin{equation}
%  \int_{p_1}^{p_2}\sqrt{\frac{\left(r'\right)^2}{1-2M/r} + r^2}d\phi=
  \int_{p_1}^{p_2}\left(\frac{\left(r'\right)^2}{1-2GM/r} + r^2\right)^\frac{1}{2}d\phi=
  \int_{p_1}^{p_2}F\left(r,r'\right)d\phi
\end{equation}

@@

% \begin{equation}\label{rdashdash_properunits}
%   r''\left(\phi\right) =
%   (r-1) + \frac{(4r-3)\left(r'\right)^2}{2r\left(r-1\right)},\qquad r>1.
% \end{equation}
\begin{equation}\label{rdashdash}
  r''\left(\phi\right) =
  (r-2GM) + \frac{(4r-6GM)\left(r'\right)^2}{2r\left(r-2GM\right)},\qquad r>2GM.
\end{equation}

\noindent 
Solutions to equation~\ref{rdashdash} correspond to a taut string
confined to Flamm's paraboloid~\cite{flamm1916}; see
Figure~\ref{flamm}.  It is interesting to compare
equation~\ref{rdashdash} with the corresponding equation for null
geodesics, which is usually given~\cite{wald} in terms of $u=2GM/r$ as
$u''=3u^2/2-u$, but is equivalent
to~$r''\left(\phi\right)=r-3GM+4GMr'/r$.

There does not appear to be a simple analytical solution to
equation~\ref{rdashdash} but nevertheless we may make several
observations.  Equation~\ref{rdashdash} is a second-order nonlinear
ordinary differential equation and thus has two constants of
integration.  There are two qualitatively different types of
solutions: those with a tangential point ($r'=0,r>2GM$), and those
without.  Consider first solutions with such a point: there exists a
pair~$\left(r,\phi\right), r>2GM$ at which~$r'(\phi)=0$.  This point
must be unique, as~$r''(\phi)>0$, and be the closest approach of the
string to the event horizon.

@@

\begin{equation}\label{roc}
  R = \frac{
    \left(r^2 + \left(r'\right)^2\right)^\frac{3}{2}
  }{
%    \left|
    r^2 + 2\left(r'\right)^2-rr''
%    \right|
  }
  =
  \frac{
    \left(r^2 + \left(r'\right)^2\right)^\frac{3}{2}
  }{
    r-\frac{\left(r'\right)^2}{2\left(r-2GM\right)}
  }\qquad\mbox{along a taut string}
\end{equation}

\noindent which would give~$R$ as a function of~$\left(r,r'\right)$.
Considering~$r'(\phi)=0$ shows that if a string is tangential at a
point with~$r_0>2GM$, then it has a radius of curvature of~$r_0^2/2GM$
at that point.  Any circular path, including a circular orbit
at~$r=r_0$, has a radius of curvature of $r_0<r_0^2/2GM$, showing that
circular orbits cannot follow a taut string at closest approach.  Also
observe that null geodesics have a radius of curvature of $r_0^2/3GM <
r_0^2/2GM$, showing that light cannot follow a taut string at closest
approach either.  The string has a point of inflection
if~$\left(r'\right)^2=2r(r-2GM)$.

@@

If, conversely, $r'(\phi)\neq 0$ whenever~$r>2GM$, then the string has
only one free end; the other is tangential to the event horizon (in
the sense that ${\displaystyle \lim_{r\longrightarrow 1^+}r'=0}$),
except in the degenerate case of constant~$\phi$ when the string is
radial for~$r>2GM$.

@@

Here I solve equation~\ref{rdashdash} using the \verb+deSolve+ package
for ordinary differential equations~\cite{soetart2010}.  Non-radial
solutions fall into two distinct classes: those with $r>2GM$ at all
points, and those which attain $r=2GM$, in which case the string is
tangential to the event horizon.

We first consider strings for which $r>2GM$ everywhere (that is,
completely exterior strings).
Figure~\ref{closest_approach_non_self_intersecting} shows a sequence
of non-self-intersecting numerical solutions to
equation~\ref{rdashdash}, rotated so that the tangential point of
closest approach occurs at~$\phi=0$.
Figure~\ref{closest_approach_self_intersecting} shows the
corresponding diagram for self-interesecting strings
(Figure~\ref{light_closest_approach} shows equivalent null geodesics
for comparison).  A bisection technique reveals that if the closest
approach is greater than a critical value of about~$\simeq 1.0761\cdot
2GM$, the string does not self-intersect; if the closest approach is
less than this value, then the string crosses itself at~$\phi=\pi/2$.
As discussed above, if $\left(r'\right)^2=2r(r-2GM)$, the string has a
point of inflection at which point the radius of curvature becomes
infinite.  Figure~\ref{curvature_switch} shows non-self intersecting
strings rotated so that the point of inflection occurs at~$\phi=0$; we
see that each string has a finite length in which it is convex to the
origin and two semi-infinite stretches in which the curvature is
negative.

There are other ways to parametrize completely exterior strings;
Figure \ref{strings_r_equals_2}, for example, shows strings passing
through~$(4GM,0)$.  Figure~\ref{light_r_equals_2} shows light paths for
comparison.

The cases where the string attains $r=2GM$ are tangent to the event
horizon (or, exceptionally, purely radial).
Figure~\ref{fixed_EH_intersection} shows strings tangential to the
event horizon at~$\phi=0$ and Figure~\ref{one_free_end_r_equals_2}
shows tangential strings passing through $(4GM,0)$.

@@

\begin{figure}[p] % created by "closest_approach2.R"
\centering
\includegraphics[width=170mm]{closest_approach_selfintersecting.pdf}
\caption{Self-intersecting strings shown for $\phi\geq 0$.  Closest
  approach to the event horizon occurs at $\phi=0$ (3 o'clock) and
  $r/2GM\in\left(1,1.0761\right)$; strings are symmetrical about $\phi=0$
  but for clarity only $\phi\geq 0$ is shown}
\label{closest_approach_self_intersecting}
\end{figure}

@@

\begin{figure}[p]
\centering
\includegraphics{angle_at_r_equals_2.pdf} % created by "angle_at_r_equals_2.R"
\caption{Strings passing through $(4GM,0)$}
\label{strings_r_equals_2}
\end{figure}

\begin{figure}[p]
\centering
\includegraphics{light_start_at_r_equals_2.pdf} % created by "angle_start_at_r_equals_2.R"
\caption{Null geodesics passing through $(4GM,0)$, shown for $0\leq\phi\leq 2\pi$}
\label{light_r_equals_2}
\end{figure}

@@

\begin{figure}[p] % created by "one_free_end_r_equals_2.R"
\centering
\includegraphics[width=170mm]{one_free_end_r_equals_2.pdf}
\caption{Strings tangential to the event horizon passing through $(4GM,0)$}
\label{one_free_end_r_equals_2}
\end{figure}

>>>>

\begin{abstract}
Here I analyze the behaviour of stationary, light inextensible strings
under tension in the vicinity of a nonrotating black hole.  Such
strings minimize their proper length, given by integrating the spatial
component of the Schwarzschild metric along the string.  The path is
calculated by numerical solution to the Euler-Lagrange equations.  If
the string is entirely outside the event horizon, its closest approach
to the singularity is tangential; if the distance of closest
approach is less than about $1.0761\cdot 2M$, the
string self-intersects.  If the string attains $r=2M$, then it
will in general be tangential to the event horizon.  Light taut
strings interpret the concept ``straight line'' and their behaviour
suggests a possible way of manipulating a black hole.
\end{abstract}
% AJP requires an abstract for all regular article submissions.
% Abstracts are optional for submissions to the "Notes and Discussions" section.

@@

\label{intro}
Light inextensible strings are a stalwart feature of elementary
physics.  A light string is not subject to gravity, and being under
tension means that its configuration is an extremum of proper length.
Of course, no real string is perfectly weightless or inextensible.
With tension~$T$, gravitational acceleration $g$ and (proper) mass per
(proper) unit length~$\mu$, then a string will behave as though light
on lengthscales $L\lesssim T/\left(\mu g\right)$.  Using Newtonian
expressions for the strength of gravity at the event horizon of a
black hole gives $L\lesssim TM/\left(\mu c^2\right)$.  For the string
to behave as though light on lengthscales comparable with the
Schwarzschild radius would require $T\gtrsim\mu c^2$.  Exceeding this
criterion would violate the dominant energy condition (and in this
context we note that transverse waves would propagate at
speed~$\sqrt{T/\mu}\sim c$).  Nevertheless, the idealization of a
string as light and inextensible is an interesting and informative
limit: its analysis interprets the concept ``straight line'' and
suggests a possible way of manipulating a black hole.  Peculiar
counterintuitive behaviour such as inward-directed centrifugal force
in the vicinity of a black hole is well
known~\citep{abramowicz1992,abramowicz1993}, and the analysis of taut
strings furnishes another everyday window into general relativity.
Another possibility, presented by \citet{damour1978} and discussed by
\citet{thorne1986}, is to pass electric current through such taut
strings, an option not pursued here.

@@

\begin{equation}\label{schwarzschild}
ds^2= -dt^2\left(1-2M/r\right) +\frac{dr^2}{1-2M/r} + r^2\left(d\theta^2 + \sin^2\theta d\phi^2\right),
\end{equation}

\noindent where units in which~$G=c=1$ are used, has been known for
over a century~\cite{schwarzschild1916}.  Here I consider the spatial
component~$ds^2= dr^2/\left(1-2M/r\right) + r^2d\phi^2$, confined to
equatorial points~($\theta=\pi/2$); we understand~$2M\leq r<\infty$
throughout.  A circular string at constant~$r$ will, trivially, have
proper length~$2\pi r$.

Considering first the radial component we may calculate the
proper length $L\left(r_1,r_2\right)$ of a string stretching
from~$r=r_1\geq 2M$ to~$r_2\geq r_1$ at constant $\theta,\phi$:

@@

\begin{equation}\label{radial_string_length}
  \int_{r=r_1}^{r_2}\frac{dr}{\sqrt{1-2M/r}}=
  \left.
  \sqrt{r(r-2M)} +\frac{1}{2}\log\left(
  \frac{\sqrt{r}+\sqrt{r-2M}}{\sqrt{r}-\sqrt{r-2M}}\right)
  \right|_{r_1}^{r_2}
  \end{equation}

Observe that integration may be performed with a lower limit
of~$r_1=2M$ with no difficulty, even though the radial component of the
Schwarzschild metric approaches infinity there.

Elementary methods show that the proper length $\ell$ of radial strings has
some peculiar features.  Firstly, consider the proper length from
$r=2M$ to $r=2M(1+w)$ where $w>0, \left|w\right|\ll 1$.  Then

@@

 \begin{equation}\label{ell}
   \ell =
   \int_{r=2M}^{r=2M(1+w)}ds
   =2M\left(2\sqrt{w} + \frac{1}{3}w^{3/2} + O(w^{5/2})\right).
 \end{equation}

This would indicate that, close to the Schwarzschild radius $r=2M$,
large (but finite) amounts of string are required to connect two
points with slightly differing radial coordinates.  Passing to the
case $w\gg 1$ we have an asymptotic expansion that begins

% \begin{equation}\label{asymptotic_ell_properunits}
%   \ell = w +  \frac{1}{2}\log w + \frac{1}{2} + \log(2)  +  \frac{1}{8}w^{-1} + O(w^{-2}),
% \end{equation}
% 
\begin{equation}\label{asymptotic_ell}
  \ell = 2M\left(w +  \frac{1}{2}\log w + \frac{1}{2} + \log(2)  +  \frac{1}{8}w^{-1} + O(w^{-2})\right),
\end{equation}

\noindent suggesting that there is ``extra radial space'', of length
over $1.19\cdot 2M$, sensible at large distances from the event
horizon.  Radial strings have proper length longer than the
Schwarzschild radial coordinate difference $w$.

@@

\begin{equation}
%  \int_{p_1}^{p_2}\sqrt{\frac{\left(r'\right)^2}{1-2M/r} + r^2}d\phi=
  \int_{p_1}^{p_2}\left(\frac{\left(r'\right)^2}{1-2M/r} + r^2\right)^\frac{1}{2}d\phi=
  \int_{p_1}^{p_2}F\left(r,r'\right)d\phi
\end{equation}

@@

% \begin{equation}\label{rdashdash_properunits}
%   r''\left(\phi\right) =
%   (r-1) + \frac{(4r-3)\left(r'\right)^2}{2r\left(r-1\right)},\qquad r>1.
% \end{equation}
\begin{equation}\label{rdashdash}
  r''\left(\phi\right) =
  (r-2M) + \frac{(4r-6M)\left(r'\right)^2}{2r\left(r-2M\right)},\qquad r>2M.
\end{equation}

\noindent 
Solutions to equation~\ref{rdashdash} correspond to a taut string
confined to Flamm's paraboloid~\cite{flamm1916}; see
Figure~\ref{flamm}.  It is interesting to compare
equation~\ref{rdashdash} with the corresponding equation for null
geodesics, which is usually given~\cite{wald} in terms of $u=2M/r$ as
$u''=3u^2/2-u$, but is equivalent
to~$r''\left(\phi\right)=r-3M+4Mr'/r$.

There does not appear to be a simple analytical solution to
equation~\ref{rdashdash} but nevertheless we may make several
observations.  Equation~\ref{rdashdash} is a second-order nonlinear
ordinary differential equation and thus has two constants of
integration.  There are two qualitatively different types of
solutions: those with a tangential point ($r'=0,r>2M$), and those
without.  Consider first solutions with such a point: there exists a
pair~$\left(r,\phi\right), r>2M$ at which~$r'(\phi)=0$.  This point
must be unique, as~$r''(\phi)>0$, and be the closest approach of the
string to the event horizon.

@@

\begin{equation}\label{roc}
  R = \frac{
    \left(r^2 + \left(r'\right)^2\right)^\frac{3}{2}
  }{
%    \left|
    r^2 + 2\left(r'\right)^2-rr''
%    \right|
  }
  =
  \frac{
    \left(r^2 + \left(r'\right)^2\right)^\frac{3}{2}
  }{
    r-\frac{\left(r'\right)^2}{2\left(r-2M\right)}
  }\qquad\mbox{along a taut string}
\end{equation}

\noindent which would give~$R$ as a function of~$\left(r,r'\right)$.
Considering~$r'(\phi)=0$ shows that if a string is tangential at a
point with~$r_0>2M$, then it has a radius of curvature of~$r_0^2/2M$
at that point.  Any circular path, including a circular orbit
at~$r=r_0$, has a radius of curvature of $r_0<r_0^2/2M$, showing that
circular orbits cannot follow a taut string at closest approach.  Also
observe that null geodesics have a radius of curvature of $r_0^2/3M <
r_0^2/2M$, showing that light cannot follow a taut string at closest
approach either.  The string has a point of inflection
if~$\left(r'\right)^2=2r(r-2M)$.

@@

If, conversely, $r'(\phi)\neq 0$ whenever~$r>2M$, then the string has
only one free end; the other is tangential to the event horizon (in
the sense that ${\displaystyle \lim_{r\longrightarrow 1^+}r'=0}$),
except in the degenerate case of constant~$\phi$ when the string is
radial for~$r>2M$.

@@

Here I solve equation~\ref{rdashdash} using the \verb+deSolve+ package
for ordinary differential equations~\cite{soetart2010}.  Non-radial
solutions fall into two distinct classes: those with $r>2M$ at all
points, and those which attain $r=2M$, in which case the string is
tangential to the event horizon.

We first consider strings for which $r>2M$ everywhere (that is,
completely exterior strings).
Figure~\ref{closest_approach_non_self_intersecting} shows a sequence
of non-self-intersecting numerical solutions to
equation~\ref{rdashdash}, rotated so that the tangential point of
closest approach occurs at~$\phi=0$.
Figure~\ref{closest_approach_self_intersecting} shows the
corresponding diagram for self-interesecting strings
(Figure~\ref{light_closest_approach} shows equivalent null geodesics
for comparison).  A bisection technique reveals that if the closest
approach is greater than a critical value of about~$\simeq 1.0761\cdot
2M$, the string does not self-intersect; if the closest approach is
less than this value, then the string crosses itself at~$\phi=\pi/2$.
As discussed above, if $\left(r'\right)^2=2r(r-2M)$, the string has a
point of inflection at which point the radius of curvature becomes
infinite.  Figure~\ref{curvature_switch} shows non-self intersecting
strings rotated so that the point of inflection occurs at~$\phi=0$; we
see that each string has a finite length in which it is convex to the
origin and two semi-infinite stretches in which the curvature is
negative.

There are other ways to parametrize completely exterior strings;
Figure \ref{strings_r_equals_2}, for example, shows strings passing
through~$(4M,0)$.  Figure~\ref{light_r_equals_2} shows light paths for
comparison.

The cases where the string attains $r=2M$ are tangent to the event
horizon (or, exceptionally, purely radial).
Figure~\ref{fixed_EH_intersection} shows strings tangential to the
event horizon at~$\phi=0$ and Figure~\ref{one_free_end_r_equals_2}
shows tangential strings passing through $(4M,0)$.

@@

\begin{figure}[p] % created by "closest_approach2.R"
\centering
\includegraphics[width=170mm]{closest_approach_selfintersecting.pdf}
\caption{Self-intersecting strings shown for $\phi\geq 0$.  Closest
  approach to the event horizon occurs at $\phi=0$ (3 o'clock) and
  $r/2M\in\left(1,1.0761\right)$; strings are symmetrical about $\phi=0$
  but for clarity only $\phi\geq 0$ is shown}
\label{closest_approach_self_intersecting}
\end{figure}

@@

\begin{figure}[p]
\centering
\includegraphics{angle_at_r_equals_2.pdf} % created by "angle_at_r_equals_2.R"
\caption{Strings passing through $(4M,0)$}
\label{strings_r_equals_2}
\end{figure}

\begin{figure}[p]
\centering
\includegraphics{light_start_at_r_equals_2.pdf} % created by "angle_start_at_r_equals_2.R"
\caption{Null geodesics passing through $(4M,0)$, shown for $0\leq\phi\leq 2\pi$}
\label{light_r_equals_2}
\end{figure}

@@

\begin{figure}[p] % created by "one_free_end_r_equals_2.R"
\centering
\includegraphics[width=170mm]{one_free_end_r_equals_2.pdf}
\caption{Strings tangential to the event horizon passing through $(4M,0)$}
\label{one_free_end_r_equals_2}
\end{figure}
